---
title: "ACT weather analysis"
author: "Muntasir Adnan"
date: "11/10/2022"
output: html_document
---

# Load the dataset
```{r}
library(tidyverse)
df <- read.csv("act_weather.csv")
head(df)
```
# remove the first column
```{r}
df <- df[-c(1)]
head(df)
```
# create a function that takes the column name as a parametre and prints minimum, median, mean and maximum
```{r}
getSummary <- function(colName)
{
  writeLines(paste(paste(colName, "summary:"),
                   "================",
                   paste("minimum : ", min(df[colName])),
                   paste("median : ", median(df[[colName]])),
                   paste("mean : ", mean(df[[colName]])),
                   paste("maximum : ", max(df[colName])),
                   sep = '\n'))
  }
```
# make a vector to hold the column names, use make.names to convert them to correct name
```{r}
targetColumns <- c("Minimum_temperature",
                  "9am_Temperature",
                  "Speed_of_maximum_wind_gust_(km/h)",
                  "3pm_Temperature")
targetColumns <- make.names(targetColumns)
```
# run a loop to get the summary
```{r}
for (columnName in targetColumns)
{
  getSummary(columnName)
}
```
# average minimum temperature of each month
```{r}
averageMinTemp <- df %>% 
  group_by(Month, Year) %>%
  summarise(avgMinTempByMonth = mean(Minimum_temperature))
head(averageMinTemp)
```
# average of the speed of maximum wind gust per each group of direction
```{r}
averageWindGust <- df %>% 
  group_by(Direction_of_maximum_wind_gust) %>% 
  summarise(mean(Speed_of_maximum_wind_gust_.km.h.))
head(averageWindGust)
```
# least amount of rainfall or dry month
```{r}
dryMonth <- df %>% 
  group_by(Month, Year) %>% 
  summarise(sumRainfall = sum(Rainfall_.mm.))
dryMonth[which.min(dryMonth$sumRainfall),]
```
# function to calculate average humidity
```{r}
avgHumidity <- function(a, b)
{
  (a+b)/2
}
```

# highest hiumidity in 2019
```{r}
highestHumid2019 <- df %>% 
  filter(Year == 2019) %>% 
  group_by(Month) %>% 
  summarise(maxHumid = max(avgHumidity(X9am_relative_humidity_..., X3pm_relative_humidity_...)))
highestHumid2019[which.max(highestHumid2019$maxHumid),]
```
# August 2020 to July 2021 summary table showing min, max and ave temp, wind speed, humidity for each month, for each quarter
```{r}
#get the subset
data20to21 <- df %>% 
  filter((Year == 2020 & Month >= 8)|(Year == 2021 & Month <= 7))

#make numeric
data20to21$X9am_wind_speed_.km.h. <- as.numeric(data20to21$X9am_wind_speed_.km.h.)
data20to21$X3pm_wind_speed_.km.h. <- as.numeric(data20to21$X3pm_wind_speed_.km.h.)

#fix NAs with mean
data20to21$X9am_wind_speed_.km.h.[is.na(data20to21$X9am_wind_speed_.km.h.)] <- mean(data20to21$X9am_wind_speed_.km.h., 
                                                                                    na.rm = TRUE)
data20to21$X3pm_wind_speed_.km.h.[is.na(data20to21$X3pm_wind_speed_.km.h.)] <- mean(data20to21$X3pm_wind_speed_.km.h., 
                                                                                    na.rm = TRUE)

#create a new data set
result <- data20to21 %>% 
  group_by(Month, Year) %>% 
  summarise(minTemp = min(Minimum_temperature), 
            maxTemp = max(Maximum_temperature), 
            aveTemp = mean((Maximum_temperature +Minimum_temperature)/2),
            minWind = min(Speed_of_maximum_wind_gust_.km.h.), 
            maxWind = max(Speed_of_maximum_wind_gust_.km.h.), 
            avWindSpeed = mean((X9am_wind_speed_.km.h. + X3pm_wind_speed_.km.h.)/2),
            avHumidity = mean((X3pm_relative_humidity_... + X9am_relative_humidity_...)/2),
            maxHumid = if (max(X3pm_relative_humidity_...)> max(X9am_relative_humidity_...)) max(X3pm_relative_humidity_...) else max(X9am_relative_humidity_...),
            minHumid = if (min(X3pm_relative_humidity_...)< min(X9am_relative_humidity_...)) min(X3pm_relative_humidity_...) else min(X9am_relative_humidity_...))

#quarter
resultQuarter <- data.frame(matrix(ncol = 9, nrow = 0))
colnames(resultQuarter) <- c("minTemp", "maxTemp", "avTemp", "minWind", "maxWind", "avWind", "minHum", "maxHum", "avHum")
for (i in c(1,4,7,10))
{
  temp <- c()
  
  temp <- append(temp, min(c(result$minTemp[i], result$minTemp[i+1], result$minTemp[i+2])))
  temp <- append(temp, max(c(result$maxTemp[i], result$maxTemp[i+1], result$maxTemp[i+2])))
  temp <- append(temp, mean(c(result$aveTemp[i], result$aveTemp[i+1], result$aveTemp[i+2])))
  
  temp <- append(temp, min(c(result$minWind[i], result$minWind[i+1], result$minWind[i+2])))
  temp <- append(temp, max(c(result$maxWind[i], result$maxWind[i+1], result$maxWind[i+2])))
  temp <- append(temp, mean(c(result$avWindSpeed[i], result$avWindSpeed[i+1], result$avWindSpeed[i+2])))
  
  temp <- append(temp, min(c(result$minHumid[i], result$minHumid[i+1], result$minHumid[i+2])))
  temp <- append(temp, max(c(result$maxHumid[i], result$maxHumid[i+1], result$maxHumid[i+2])))
  temp <- append(temp, mean(c(result$avHumidity[i], result$avHumidity[i+1], result$avHumidity[i+2])))
  
  resultQuarter[nrow(resultQuarter) + 1,] <- temp
  
}
```
# tests
```{r}
sum(is.na(data20to21$X3pm_relative_humidity_...))
sum(is.na(data20to21$X9am_relative_humidity_...))
result$minTemp[2]
min(c(result$minTemp[1], result$minTemp[1+1], result$minTemp[1+2]))
```

